---
title: "An Exploration of Health Through Air Quality Measures and Substance Abuse"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

    -   The data is from the Center for Disease Control (CDC )and it is "Air Quality Measures on the National Environmental Health Tracking Network"

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   The data was originally collected by the Environmental Health Tracking Network. This data which provides values about air pollution was collected from Air Quality Systems (AQS) from 4,000 monitoring stations in the years 2001 - 2011.

-   Write a brief description of the observations.

    -   The observations

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    How do air pollution levels differ between states from the North, South, East, and West of the mainland United States?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-1
air_quality <- read_csv("data/Air_Quality_Measures_on_the_National_Environmental_Health_Tracking_Network (1).csv")
 
glimpse(air_quality)

```

# Data 2

## Introduction and data

-   Identify the source of the data.

    -   The population data is from the World Census and the data about death risk factors was acquired from the World Health Organization.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   The World census estimates population based on census, survey, and administrative information. The World Health Organization death risk factors data was reported by member states from Civil Registration and Vital Statistics, which records births, deaths, and causes of death.

-   Write a brief description of the observations.

    -   Each row represents one country in one year and each column represents a cause of death, such as from nutrition, alcochol/drug abuse, and air pollution, among other things.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    -   What health risk factors (like diet, health, physical activity, etc.) have the greatest association with substance abuse death among the largest countries in population?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    -   The research topic is about how different health factors are related to substance abuse. We hypothesize that greater substance abuse deaths would be related to more deaths from nutrition, smoking, and vitamin deficiencies, representing a pattern of poor choices and habits.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   The types of variables in this question are mostly quantitative since the number of deaths from substance abuse is a continuous variable. Most of the other death risk factors are also continuous quantitative variables, although it would be possible to add a categorical variable representing continent later in the project.

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

risk_factors <- read_csv("data/number-of-deaths-by-risk-factor.csv")
population <- read_csv("data/populations.csv")

glimpse(risk_factors)
glimpse(population)

```

# Data 3

## Introduction and data

-   Identify the source of the data.

    -   The data is from the 2018 National Survey on Drug Use and Health (NSDUH).

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   This dataset was put together for the CORGIS Dataset Project by Austin Cory Bart, Ryan Whitcomb, Joung Min Choi, and Bo Guan. The data was collected from the Substance Abuse and Mental Health Data Archive (SAMHDA).

-   Write a brief description of the observations.

    -   This dataset is about substance abuse (cigarettes, marijuana, cocaine, alcohol, tobacco) among different age groups and states, separated by year. Each row represents one state in one year, with columns indicating the rates of use, total number of people who use, rates of use disorder, and total number of people who suffer from use disorder in the past year and in the past month for each drug by people in various age groups.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    -   Which state(s) in the US has/have the worst teenage substance abuse problem, and how has the situation there changed from 2002-2018?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    -   The primary research topic is teenage substance abuse, and which state suffers most from this problem.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   The two main categorical variables in this research question are State and Year. Most other variables such as total use and use rate for various drugs, as well as population, are quantitative variables.

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-3

drugs <- read_csv("data/drugs.csv")

glimpse(drugs)
```
