---
title: "Project title"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

# Introduction and Data

Our research is motivated by the desire to understand how air pollution differs between the various regions in the United States. We will be investigating the research question of how do air pollution levels differ between states from the West, Southwest, Midwest, Southeast, and Northeast as defined by the National Geographic Society of the mainland United States from the years 2008-2011?

In order to achieve our goal, we will be performing statistical analysis on the annual average ambient concentrations of PM 2.5 in micrograms per cubic meter, based on seasonal averages and daily measurement (monitor and modeled data) measurement in the air pollution data set. We hypothesize that highly Northeast will have the most air pollution as it is the larger industrial region in the Untied States. The air pollution data set is obtained from the Center for Disease Control (CDC). The data was originally collected by the Environmental Health Tracking Network. This data provides values about air pollution collected from Air Quality Systems (AQS) from 4,000 monitoring stations across the United States. It is also important to note that the data was collected in a completely ethical manner as regulated by CDC guidelines.

# Methodology

```{r}
#| label: load-packages

library(tidyverse)
library(tidymodels)

```

```{r}
#| label: load-data

air_quality <- read_csv(file = "data/Air_Quality_Measures_on_the_National_Environmental_Health_Tracking_Network (1).csv")
```

```{r}
#| label: select-statistic

air_quality <- air_quality |>
  filter(MeasureName == "Annual average ambient concentrations of PM 2.5 in micrograms per cubic meter, based on seasonal averages and daily measurement (monitor and modeled data)") |>
  select(StateName, CountyName, ReportYear, Value) |>
  rename(avg_concentration = Value)

```

```{r}
#| label: histogram

air_quality |>
  filter(ReportYear >= 2008) |>
  ggplot(aes(x = avg_concentration)) +
  geom_histogram(fill = 'steelblue', color = 'black') +
  labs(x = "Annual Average Concentration of PM 2.5 (micrograms per cubic meter",
       y = "Count",
       title = "Distribution of County Fine Particle Concentration",
       subtitle = "2008 - 2011")

```

-   Comparison: We'll compare the results of our air pollution survey with Census population data, population increase, total land area, number of firms, per capita income, metropolitan area (boolean), and population over 65

# Results
