---
title: "Project title"
subtitle: "Report"
format: html
editor: visual
execute:
  echo: true
---

# Introduction and Data

Our research is motivated by the desire to understand how air pollution differs between the various regions in the United States. We will be investigating the research question of how do air pollution levels differ between states from the West, Southwest, Midwest, Southeast, and Northeast as defined by the National Geographic Society of the mainland United States from the years 2008-2011?

In order to achieve our goal, we will be performing statistical analysis on the annual average ambient concentrations of PM 2.5 in micrograms per cubic meter, based on seasonal averages and daily measurement (monitor and modeled data) in the air pollution data set. PM 2.5 refers to particles in the air that are 2.5 microns (0.001 mm) or less in width. Because particles in the PM 2.5 size range can travel through the respiratory tract and reach the lungs, they can cause short and long term health problems. As such, PM 2.5 concentration is commonly used as a measurement for air quality and air pollution.

We hypothesize that highly Northeast will have the most air pollution as it is the largest industrial region in the Untied States. The air pollution data set is obtained from the Center for Disease Control (CDC). The data was originally collected by the Environmental Health Tracking Network. This data provides values about air pollution collected from Air Quality Systems (AQS) from 4,000 monitoring stations across the United States. It is also important to note that the data was collected in a completely ethical manner as regulated by CDC guidelines.

# Methodology

To start our analysis, we must load in the neecessary packages and data.

```{r}
#| label: load-packages-data

library(tidyverse)
library(tidymodels)

air_quality <- read_csv(file = "data/Air_Quality_Measures_on_the_National_Environmental_Health_Tracking_Network (1).csv")

```

The `air_quality` dataset contains several statistics on fine particle concentrations and ozone concentrations on the county level in the United States from 2001 to 2008. Our analysis will focus on the annual average ambient concentrations of PM 2.5 based on seasonal averages and daily measurement, and we will be looking at the data from 2008 to 2011.

```{r}
#| label: select-statistic

concentrations <- air_quality |>
  filter(MeasureName == "Annual average ambient concentrations of PM 2.5 in micrograms per cubic meter, based on seasonal averages and daily measurement (monitor and modeled data)",
         ReportYear >= 2008) |>
  select(StateName, CountyName, ReportYear, Value) |>
  rename(avg_concentration = Value)

```

The distribution of average county PM 2.5 concentration in the United States from 2008-2011 can be seen with a histogram.

```{r}
#| label: histogram

concentrations |>
  ggplot(aes(x = avg_concentration)) +
  geom_histogram(binwidth = 0.5,
                 fill = 'steelblue', 
                 color = 'black') +
  labs(x = "Annual Average Concentration of PM 2.5 (micrograms per cubic meter)",
       y = "Count",
       title = "Distribution of County Fine Particle Concentration",
       caption = "data from 'Air Quality Measures on the National Environmental Health \n Tracking Network' collected by the Environmental Protection Agency (EPA)",
       subtitle = "2008 - 2011")

```

The distribution of annual average PM 2.5 concentrations is right skewed, so there are a few counties with large outliers. In general, concentrations range between about 4 and 22 micrograms per cubic meter. The peak occurs at 10 micrograms per cubic meter, and the graph is slightly left-skewed.

```{r}
#| label: median-concentration

concentrations |>
  summarize(mean_avg_concentration = mean(avg_concentration))

```

The median annual average PM 2.5 concentration is 9.64 micrograms per cubic meter. This is the typical value for a county in the United States from 2008-2011.

We are exploring how air quality differs by geographic region as defined by the National Geographic Society. The regions include the Northeast, Southeast, Midwest, Southwest, and West.

```{r}
#| label: state-regions

northeast <- c("Maine", "Vermont", "New Hampshire", "Massachusetts", 
               "Rhode Island", "Connecticut", "New York", "New Jersey",
               "Pennsylvania")

southeast <- c("Delaware", "Maryland", "West Virginia", "Virginia", 
               "North Carolina", "District of Columbia", "Kentucky",
               "Tennessee", "South Carolina", "Georgia", "Florida", 
               "Alabama", "Mississippi", "Arkansas", "Louisiana")

midwest <- c("Ohio", "Indiana", "Michigan", "Wisconsin", "Illinois",
             "Minnesota", "Iowa", "Missouri", "North Dakota", 
             "South Dakota", "Nebraska", "Kansas")

southwest <- c("Arizona", "New Mexico", "Oklahoma", "Texas")

west <- c("Montana", "Wyoming", "Colorado", "Utah", "Idaho", "Nevada",
          "Washington", "Oregon", "California", "Hawaii", "Alaska")

concentrations <- concentrations |>
  mutate(Region = case_when(
    StateName %in% northeast ~ "Northeast",
    StateName %in% southeast ~ "Southeast",
    StateName %in% midwest ~ "Midwest",
    StateName %in% southwest ~ "Southwest",
    StateName %in% west ~ "West"
    )
  )

```

Using this division of regions, the Southeast and Midwest regions have the most counties, while the Northeast region has the least. However, all regions have at least 800 counties.

```{r}
#| label: num-counties

concentrations |>
  ggplot(aes(x = fct_infreq(Region))) +
  geom_bar(fill = 'steelblue', color = 'black') +
  labs(x = "Region",
       y = "Number of Counties in Dataset",
       caption = "data from 'Air Quality Measures on the National Environmental Health \n Tracking Network' collected by the Environmental Protection Agency (EPA)",
       title = "Counties by Region")

```

We compared the median annual concentration of fine particles for counties in each region to get an initial look of the regions suffering from poor air quality the most.

```{r}
#| label: compare-aq

concentrations |>
  group_by(Region) |>
  summarize(median_concen = median(avg_concentration)) |>
  arrange(-median_concen)

```

The Southeast region has the greatest median concentration of PM 2.5 particles (10.6 micrograms per cubic meter), while the West region has the lowest (6.97 micrograms per cubic meter).

```{r}

concentrations |>
  ggplot(aes(x = avg_concentration, y = Region)) +
  geom_boxplot(fill = 'steelblue') +
  labs(x = "Annual Average Concentration of PM 2.5 (micrograms per cubic meter)",
       y = "Region",
       caption = "data from 'Air Quality Measures on the National Environmental Health \n Tracking Network' collected by the Environmental Protection Agency (EPA)",
       title = "County Fine Particle Concentration by Region")

```

Looking at the distributions of average PM 2.5 concentration by region, it is evident that there are many counties in the West region with high average PM 2.5 concentrations. In fact, several of the counties with the greatest average PM 2.5 concentrations were from the West region.

```{r}
#| label: northeast-interval
#| eval: FALSE

set.seed(123)

boot_df <- concentrations |>
  filter(region == "") |>
  specify(response = avg_concentration) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "median")

boot_df |>
  summarize(lower = quantile(stat, 0.025),
            upper = quantile(stat, 0.975))
```

-   Comparison: We'll compare the results of our air pollution survey with Census population data, population increase, total land area, number of firms, per capita income, metropolitan area (boolean), and population over 65

# Results

The data analysis and visualization techniques we used included creating a histogram was created to make meaning of the mean annual average concentration of fine particle concentration calculated to ensure the mean calculated is significant for most data points which is true as the data peaks near the 9.64 mean value calculated. We also created a counties by region graph to understand the statistical significance of the concentration data calculated for each region. Since the number of counties is the greatest for the Southeast and the Midwest region, the results of these two regions resembles have the highest chance of resembling closest to reality making them more statistically significant than the regions Southwest, West and Northeast.

Our data analysis allowed us to find that overall the median annual average PM 2.5 concentration was 9.64 micrograms per cubic meter. Then, the median annual average PM 2.5 concentration was calculated for each region. Southeast had the largest median annual average PM 2.5 concentration with a value of 10.6 micrograms per cubic meter followed by the Midwest with 9.6 micrograms per cubic meter, then Northeast with 9.3 micrograms per cubic meter, Southwest with 8.3 micrograms per cubic meter, and West with 7.0 micrograms per cubic meter. A visualization of a histogram visually displays how the median differs between the regions. It is important to note that several of the regions have outliers especially the West region.

Based on the visualizations and summary statistics, we came to the conclusion that the Southeast had the highest median annual average PM 2.5 concentration while the West had the lowest median annual average PM 2.5 concentration compared to the five main regions within the United States.

# Bibliography

Centers for Disease Control and Prevention. (2018, June 5). *Air quality measures on the National Environmental Health Tracking Network.* <https://data.cdc.gov/Environmental-Health-Toxicology/Air-Quality-Measures-on-the-National-Environmental/cjae-szjv>

New York State Department of Health. (2018, February). *Fine particles (PM 2.5) questions and answers.* <https://www.health.ny.gov/environmental/indoors/air/pmq_a.htm#:~:text=The%20term%20fine%20particles%2C%20or,microns%20or%20less%20in%20width>.

OpenIntro. (n.d.). *United States counties.* <https://www.openintro.org/data/?data=county_complete>

*United States regions.* (n.d.). National Geographic. Retrieved April 4, 2023, from <https://education.nationalgeographic.org/resource/united-states-regions/>

\

